### 2021-07-19 manuscript.Rmd
More model updates per review with MJZ.

---
title: "Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment"
author: "Brendan J. Kelly, MD, MS"
date: "7/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2021-07-19 scripts/model_distance.R

@@ -189,6 +189,90 @@ p_scv2_linear_distance_site_category
 #' - nested random effects? for high-touch
 #' ###################################
 
+#' ####################################################
+#' generative model for SARS-CoV-2 contamination
+#' - linear abundance data
+#' - binomial probability of detection
+#' ####################################################
+
+dat_long %>%
+  filter(!site_descriptor %in% c("sink","toilet","wall_bathroom")) %>%
+  mutate(scv2_detected = !is.na(copies_max),
+         site_level = case_when(grepl("floor", site_descriptor) ~ "Floor",
+                                grepl("floor", site_descriptor) == FALSE ~ "Elevated")) %>% 
+  #count(site_descriptor, site_level) 
+  select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_level, touch, distance, copies_max, scv2_detected) %>%
+  distinct() %>%
+  identity() -> dat
+
+dat
+
+
+#' run binomial model
+dat %>%
+  filter(site_level != "Bathroom") %>%
+  brm(formula = scv2_detected ~ 0 + (1 + distance | site_level / touch),
+      data = .,
+      family = bernoulli,
+      chains = 4,
+      cores = 4,
+      control = list("adapt_delta" = 0.999, max_treedepth = 18),
+      backend = "cmdstanr",
+      seed = 16) -> m_binom_scv2_distance_mix_level
+
+m_binom_scv2_distance_mix_level %>% write_rds(file = "./models/binomial/m_binom_scv2_distance_mix_level.rds.gz", compress = "gz")
+m_binom_scv2_distance_mix_level <- read_rds(file = "./models/binomial/m_binom_scv2_distance_mix_level.rds.gz")
+
+m_binom_scv2_distance_mix_level
+rstan::check_hmc_diagnostics(m_binom_scv2_distance_mix_level$fit)
+m_binom_scv2_distance_mix_level %>% pp_check()
+
+m_binom_scv2_distance_mix_level %>%
+  posterior_summary() %>%
+  as_tibble(rownames = "param") %>%
+  gt::gt() %>%
+  gt::fmt_number(columns = 2:5, n_sigfig = 3)
+
+
+#' fitted
+m_binom_scv2_distance_mix_level$data %>%
+  as_tibble() %>%
+  expand(distance = modelr::seq_range(distance, n = 100),
+         site_level = unique(site_level),
[…]

### 2021-07-25 scripts/model_distance.R
Re-run models.

@@ -26,18 +26,20 @@ dat_long %>%
 
 #' ####################################################
 #' generative model for SARS-CoV-2 contamination
-#' - linear abundance data
+#' - initial sample only
+#' - elevated vs floor (no accounting for high-touch)
 #' - binomial probability of detection
 #' ####################################################
 
-dat_long %>%
+dat_first %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
   mutate(scv2_detected = !is.na(copies_max),
-         site_category = case_when(grepl("wall", site_descriptor) ~ "wall",
-                                   grepl("floor", site_descriptor) ~ "floor",
-                                   grepl("sink|toilet", site_descriptor) ~ "bathroom",
-                                   grepl("doorknob|bed|keyboard|mouse", site_descriptor) ~ "high_touch")) %>%
-  #count(site_descriptor, site_category)
-  select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_category, touch, distance, copies_max, scv2_detected) %>%
+         site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
+                                   grepl("floor",site_descriptor) == FALSE ~ "elevated"),
+         high_touch = touch == "High") %>%
+  #count(site_descriptor, site_category, high_touch)
+  select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_category, touch, high_touch, distance, copies_max, scv2_detected) %>%
   mutate(site_category = stringr::str_to_title(gsub("_"," ",site_category))) %>%
   distinct() %>%
   identity() -> dat
@@ -56,13 +58,12 @@ dat %>%
 
 #' run binomial model
 dat %>%
-  filter(site_category != "Bathroom") %>%
-  brm(formula = scv2_detected ~ 0 + (1 + distance | site_category),
+  brm(formula = log10(copies_max) ~ 0 + (1 + distance | site_category),
       data = .,
       family = bernoulli,
       chains = 4,
       cores = 4,
-      control = list("adapt_delta" = 0.999, max_treedepth = 18),
[…]

### 2021-07-25 scripts/model_time.R

@@ -31,13 +31,14 @@ dat_long %>%
 #' ####################################################
 
 dat_long %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
   mutate(scv2_detected = !is.na(copies_max),
-         site_category = case_when(grepl("wall", site_descriptor) ~ "wall",
-                                   grepl("floor", site_descriptor) ~ "floor",
-                                   grepl("sink|toilet", site_descriptor) ~ "bathroom",
-                                   grepl("doorknob|bed|keyboard|mouse", site_descriptor) ~ "high_touch")) %>%
-  #count(site_descriptor, site_category)
-  select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_category, touch, distance, copies_max, scv2_detected) %>%
+         site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
+                                   grepl("floor",site_descriptor) == FALSE ~ "elevated"),
+         high_touch = touch == "High") %>%
+  #count(site_descriptor, site_category, high_touch)
+  select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_category, touch, high_touch, distance, copies_max, scv2_detected) %>%
   mutate(site_category = stringr::str_to_title(gsub("_"," ",site_category))) %>%
   distinct() %>%
   identity() -> dat
@@ -55,18 +56,18 @@ dat %>%
 
 
 #' run binomial model
-# dat %>%
-#   filter(subject_covid_day >= 0 & subject_covid_day < 10) %>%
-#   brm(formula = scv2_detected ~ 0 + (1 + subject_covid_day | site_category),
-#       data = .,
-#       family = bernoulli,
-#       chains = 4,
-#       cores = 4,
-#       control = list("adapt_delta" = 0.999, max_treedepth = 18),
-#       backend = "cmdstanr",
-#       seed = 16) -> m_binom_scv2_time_mix_category
-# 
-# m_binom_scv2_time_mix_category %>% write_rds(file = "./models/binomial/m_binom_scv2_time_mix_category.rds.gz", compress = "gz")
+dat %>%
[…]

### 2021-07-26 manuscript.Rmd
Add distance hurdle model.

@@
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-## R Markdown
-
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
-```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2021-07-26 scripts/model_distance.R

@@ -15,13 +15,14 @@ set.seed(16)
 read_csv("./data/dat_long.csv") %>%
   #mutate(unit = replace(unit, unit=="M1", "R1")) %>%
   identity() -> dat_long
+dat_long
 
 dat_long %>%
   group_by(redcap_id) %>%
   filter(subject_study_day == min(subject_study_day)) %>%
   ungroup() %>%
   identity() -> dat_first
-
+dat_first
 
 
 #' ####################################################
@@ -58,7 +59,8 @@ dat %>%
 
 #' run binomial model
 dat %>%
-  brm(formula = log10(copies_max) ~ 0 + (1 + distance | site_category),
+  select(scv2_detected, distance, site_category) %>%
+  brm(formula = scv2_detected ~ 0 + (1 + distance | site_category),
       data = .,
       family = bernoulli,
       chains = 4,
@@ -144,6 +146,7 @@ dat %>%
 
 #' run binomial model
 dat %>%
+  select(scv2_detected, distance, site_category, high_touch) %>%
   brm(formula = scv2_detected ~ 0 + (1 + distance | site_category / high_touch),
       data = .,
       family = bernoulli,
@@ -247,8 +250,9 @@ dat %>%
   gt::gt()
 
 
-#' run linear model
+#' run linear model: does NOT account for specimens with unmeasured copy numbers
 dat %>%
+  select(copies_max, distance, site_category) %>%
   brm(formula = log10(copies_max) ~ 0 + (1 + distance | site_category),
       data = .,
       family = gaussian,
@@ -289,8 +293,8 @@ m_linear_scv2_distance_mix_category_fitted %>%
   #mutate(high_touch = case_when(high_touch == TRUE ~ "High Touch",
   #                              high_touch == FALSE ~ "Low Touch")) %>%
   ggplot(aes(x = distance, y = .value)) +
-  geom_point(data = m_linear_scv2_distance_mix_category$data, aes(x = distance, y = log10(copies_max)), color = "grey", alpha = 0.5) +
   stat_lineribbon() +
+  geom_point(data = m_linear_scv2_distance_mix_category$data, aes(x = distance, y = log10(copies_max)), color = "grey", alpha = 0.5) +
   #facet_wrap(facets = ~ site_category + high_touch, scales = "fixed") +
   facet_wrap(facets = ~ site_category) +
   scale_fill_brewer(palette = "Reds") +
@@ -354,12 +358,13 @@ dat %>%
 
 #' run linear model
 dat %>%
[…]

### 2021-07-27 scripts/model_distance.R
Hurdle models for time from COVID-19 diagnosis.

@@ -10,7 +10,7 @@ set.seed(16)
 
 
 #' ####################################################
-#' read data and generate descriptive tables and plots
+#' read data
 #' ####################################################
 read_csv("./data/dat_long.csv") %>%
   #mutate(unit = replace(unit, unit=="M1", "R1")) %>%

### 2021-07-27 scripts/model_time.R

@@ -10,23 +10,25 @@ set.seed(16)
 
 
 #' ####################################################
-#' read data and generate descriptive tables and plots
+#' read data
 #' ####################################################
 read_csv("./data/dat_long.csv") %>%
   #mutate(unit = replace(unit, unit=="M1", "R1")) %>%
   identity() -> dat_long
+dat_long
 
 dat_long %>%
   group_by(redcap_id) %>%
   filter(subject_study_day == min(subject_study_day)) %>%
   ungroup() %>%
   identity() -> dat_first
-
+dat_first
 
 
 #' ####################################################
 #' generative model for SARS-CoV-2 contamination
-#' - linear abundance data
+#' - initial sample only
+#' - elevated vs floor (no accounting for high-touch)
 #' - binomial probability of detection
 #' ####################################################
 
@@ -50,20 +52,20 @@ dat
 #' get prior
 dat %>%
   brms::get_prior(data = ., family = bernoulli,
-                  scv2_detected ~ 0 + (1 + distance | site_category)
+                  scv2_detected ~ (1 + subject_covid_day | site_category)
   ) %>%
   gt::gt()
 
 
 #' run binomial model
 dat %>%
-  filter(subject_covid_day >= 0 & subject_covid_day < 10) %>%
+  select(scv2_detected, subject_covid_day, site_category) %>%
   brm(formula = scv2_detected ~ 0 + (1 + subject_covid_day | site_category),
       data = .,
       family = bernoulli,
       chains = 4,
       cores = 4,
-      control = list("adapt_delta" = 0.99999, max_treedepth = 16),
+      control = list("adapt_delta" = 0.999, max_treedepth = 10),
       backend = "cmdstanr",
       seed = 16) -> m_binom_scv2_time_mix_category
 
@@ -85,19 +87,25 @@ m_binom_scv2_time_mix_category %>%
 m_binom_scv2_time_mix_category$data %>%
   as_tibble() %>%
   expand(subject_covid_day = modelr::seq_range(subject_covid_day, n = 100),
-         site_category = unique(site_category)) %>%
+         site_category = unique(site_category)#,
+         #high_touch = unique(high_touch)
+  ) %>%
+  #filter(!(high_touch == TRUE & site_category == "Floor")) %>%
[…]

### 2021-07-29 scripts/model_distance.R
Notes on unit transfers, subject-level differences.

@@ -39,14 +39,13 @@ dat_first %>%
          site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
                                    grepl("floor",site_descriptor) == FALSE ~ "elevated"),
          high_touch = touch == "High") %>%
-  #count(site_descriptor, site_category, high_touch)
+  #count(site_descriptor, site_category, high_touch) %>% gt()
   select(redcap_id, subject_covid_day, subject_hosp_day, subject_covid_day, swab_site, unit, site_category, touch, high_touch, distance, copies_max, scv2_detected) %>%
   mutate(site_category = stringr::str_to_title(gsub("_"," ",site_category))) %>%
   distinct() %>%
   identity() -> dat
 
 dat
-
 
 
 #' get prior
@@ -58,18 +57,18 @@ dat %>%
 
 
 #' run binomial model
-dat %>%
-  select(scv2_detected, distance, site_category) %>%
-  brm(formula = scv2_detected ~ 0 + (1 + distance | site_category),
-      data = .,
-      family = bernoulli,
-      chains = 4,
-      cores = 4,
-      control = list("adapt_delta" = 0.999, max_treedepth = 10),
-      backend = "cmdstanr",
-      seed = 16) -> m_binom_scv2_distance_mix_category
-
-m_binom_scv2_distance_mix_category %>% write_rds(file = "./models/binomial/m_binom_scv2_distance_mix_category.rds.gz", compress = "gz")
+# dat %>%
+#   select(scv2_detected, distance, site_category) %>%
+#   brm(formula = scv2_detected ~ 0 + (1 + distance | site_category),
+#       data = .,
+#       family = bernoulli,
+#       chains = 4,
+#       cores = 4,
+#       control = list("adapt_delta" = 0.999, max_treedepth = 10),
+#       backend = "cmdstanr",
+#       seed = 16) -> m_binom_scv2_distance_mix_category
+# 
+# m_binom_scv2_distance_mix_category %>% write_rds(file = "./models/binomial/m_binom_scv2_distance_mix_category.rds.gz", compress = "gz")
 m_binom_scv2_distance_mix_category <- read_rds(file = "./models/binomial/m_binom_scv2_distance_mix_category.rds.gz")
 
 m_binom_scv2_distance_mix_category
@@ -145,18 +144,18 @@ dat %>%
 
 
 #' run binomial model
-dat %>%
[…]

### 2021-07-29 scripts/model_time.R

@@ -58,18 +58,18 @@ dat %>%
 
 
 #' run binomial model
-dat %>%
-  select(scv2_detected, subject_covid_day, site_category) %>%
-  brm(formula = scv2_detected ~ 0 + (1 + subject_covid_day | site_category),
-      data = .,
-      family = bernoulli,
-      chains = 4,
-      cores = 4,
-      control = list("adapt_delta" = 0.999, max_treedepth = 10),
-      backend = "cmdstanr",
-      seed = 16) -> m_binom_scv2_time_mix_category
-
-m_binom_scv2_time_mix_category %>% write_rds(file = "./models/binomial/m_binom_scv2_time_mix_category.rds.gz", compress = "gz")
+# dat %>%
+#   select(scv2_detected, subject_covid_day, site_category) %>%
+#   brm(formula = scv2_detected ~ 0 + (1 + subject_covid_day | site_category),
+#       data = .,
+#       family = bernoulli,
+#       chains = 4,
+#       cores = 4,
+#       control = list("adapt_delta" = 0.999, max_treedepth = 10),
+#       backend = "cmdstanr",
+#       seed = 16) -> m_binom_scv2_time_mix_category
+# 
+# m_binom_scv2_time_mix_category %>% write_rds(file = "./models/binomial/m_binom_scv2_time_mix_category.rds.gz", compress = "gz")
 m_binom_scv2_time_mix_category <- read_rds(file = "./models/binomial/m_binom_scv2_time_mix_category.rds.gz")
 
 m_binom_scv2_time_mix_category
@@ -145,18 +145,18 @@ dat %>%
 
 
 #' run binomial model
-dat %>%
-  select(scv2_detected, subject_covid_day, site_category, high_touch) %>%
-  brm(formula = scv2_detected ~ 0 + (1 + subject_covid_day | site_category / high_touch),
-      data = .,
-      family = bernoulli,
-      chains = 4,
-      cores = 4,
-      control = list("adapt_delta" = 0.9999, max_treedepth = 10),
-      backend = "cmdstanr",
-      seed = 16) -> m_binom_scv2_time_mix_category_touch
-
-m_binom_scv2_time_mix_category_touch %>% write_rds(file = "./models/binomial/m_binom_scv2_time_mix_category_touch.rds.gz", compress = "gz")
+# dat %>%
+#   select(scv2_detected, subject_covid_day, site_category, high_touch) %>%
[…]

### 2021-08-16 manuscript.Rmd
Add interaction model for pandemic duration.

@@
 ---
-title: "Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment"
-author: "Brendan J. Kelly, MD, MS"
-date: "7/19/2021"
+#title: "Spatial and temporal effects on SARS-CoV-2 contamination of the healthcare environment"
+#author: "Brendan J. Kelly, MD, MS"
 output: html_document
 ---
+
+# Results
+
+### Distance Effect Negligible
+
+### Decay over Time
+
+### Heterogeneity among Subjects in Probability of Detection
+
+### Total Study Day (i.e., Time Into Second Wave) Accounts for Floor Detection Probability
+
+### What Subject-Level Factors Account for Detection Among High-Touch Elevated Sites?

### 2021-08-16 scripts/model_distance.R

@@ -1074,12 +1074,13 @@ m_mvbinom_scv2_distance_category_touch_mix_subject_fitted %>%
   pivot_wider(id_cols = c(subject_room_id, numeric_id), names_from = site_category, values_from = .epred) %>%
   left_join(summarise(group_by(dat_long, subject_room_id), total_study_day = min(total_study_day, na.rm = TRUE), subject_room_day = min(subject_room_day, na.rm = TRUE), unit = unit[1]), by = "subject_room_id") %>%
   #mutate(numeric_id = replace(numeric_id, Floor <= 0.75 | Elevated <= 0.75, NA)) %>%
-  ggplot(data = ., aes(x = Floor, y = Elevated, color = subject_room_id, label = numeric_id)) +
-  #geom_point() +
-  geom_point(shape = 21, color = "black") +
+  ggplot(data = ., aes(x = Floor, y = Elevated, color = total_study_day, label = numeric_id)) +
+  geom_point() +
+  #geom_point(shape = 21, color = "black") +
   #ggrepel::geom_text_repel(color = "black", max.overlaps = 50) +
   coord_equal(xlim = c(0,1), ylim = c(0,1)) +
   #colorspace::scale_color_discrete_sequential() +
+  scale_color_viridis_c(option = "plasma") +
   labs(x = "Probability of SARS-CoV-2<br>Detection on Floor",
        y = "Probability of SARS-CoV-2<br>Detection on High-Touch Elevated",
   ) +
@@ -1089,7 +1090,7 @@ m_mvbinom_scv2_distance_category_touch_mix_subject_fitted %>%
         axis.text.y = ggtext::element_markdown(color = "black"),
         axis.title.x = ggtext::element_markdown(color = "black"),
         axis.title.y = ggtext::element_markdown(color = "black"),
-        legend.position = "none",
+        #legend.position = "none",
         #legend.background = element_rect(fill = "white", color = "black", size = 0.25),
         strip.background = element_blank())
 

### 2021-08-16 scripts/model_manuscript.R

@@ -1,0 +1,626 @@
+#' ##############################
+#' load libraries and set seed
+#' ##############################
+library(tidyverse)
+library(gt)
+library(tidybayes)
+library(brms)
+
+set.seed(16)
+
+
+#' ####################################################
+#' read data
+#' ####################################################
+read_csv("./data/dat_long.csv") %>%
+  identity() -> dat_long
+dat_long
+
+
+dat_long %>%
+  group_by(subject_room_id) %>%
+  filter(subject_room_day == min(subject_room_day)) %>%
+  ungroup() %>%
+  identity() -> dat_first
+dat_first
+
+
+
+
+#' ####################################################
+#' ####################################################
+#' SPATIAL MODELS
+#' ####################################################
+#' ####################################################
+
+dat_first %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
+  mutate(scv2_detected = !is.na(copies_max),
+         site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
+                                   grepl("floor",site_descriptor) == FALSE ~ "elevated"),
+         high_touch = touch == "High") %>%
+  #count(site_descriptor, site_category, high_touch) %>% gt()
+  select(redcap_id, subject_covid_day, subject_hosp_day, subject_room_day, total_study_day, swab_site, unit, site_category, touch, high_touch, distance, copies_max, scv2_detected) %>%
+  mutate(site_category = stringr::str_to_title(gsub("_"," ",site_category))) %>%
+  distinct() %>%
+  identity() -> dat
+
+dat
+
+
+#' ####################################################
+#' generative model for SARS-CoV-2 contamination
+#' - initial sample only
+#' - elevated vs floor -- accounting for high-touch
+#' - binomial probability of detection
+#' ####################################################
+
+dat_first %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
[…]

### 2021-08-16 manuscript.Rmd
Sketch manuscript outline.

--- 2021-07-26 manuscript.Rmd
@@
 ---
-title: "Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment"
-author: "Brendan J. Kelly, MD, MS"
-date: "7/19/2021"
+#title: "Spatial and temporal effects on SARS-CoV-2 contamination of the healthcare environment"
+#author: "Brendan J. Kelly, MD, MS"
+bibliography: bib/references.bib
+csl: csl/infection-control-and-hospital-epidemiology.csl
 output: html_document
+css: style_helvetica_double.css
 ---
+
+```{r setup, echo=FALSE, eval=TRUE, include=FALSE, autodep=TRUE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+library(tidyverse)
+library(tidybayes)
+library(gt)
+```
+
+## Title  
+Spatial and temporal effects on SARS-CoV-2 contamination of the healthcare environment
+
+## Abbreviated Title  
+SARS-CoV-2 Healthcare Contamination
+
+## Authors  
+Matthew J.
+Ziegler, MD, MS (1,2,3); Selamawit Bekele (2,3); Sean Loughrey, MS (2,3); Emily Reesey, MS (2,2); Elizabeth Huang (2,3); Pam Tolomeo, MPH (2,3); Ebbing Lautenbach, MD, MPH, MS (1,2,3); Brendan J.
+Kelly, MD, MS (1,2,3) for the CDC Prevention Epicenters Program
+
+## Affiliations  
+1 - Division of Infectious Diseases, Department of Medicine;   
+2 - Department of Biostatistics, Epidemiology, and Informatics;  
+3 - Center for Clinical Epidemiology and Biostatistics, Perelman School of Medicine, University of Pennsylvania, Philadelphia, PA
+
+## Corresponding Author Contact  
+Brendan J. Kelly, MD, MS - brendank@pennmedicine.upenn.edu
+
+## Authors' Contributions  
+MJZ - study design, data collection and analysis, manuscript;  
+SB - subject enrollment and specimen collection;  
+SL - specimen processing and analysis;  
+ER - specimen processing and analysis;  
+EH - specimen collection and processing;   
+PT - subject enrollment and specimen collection, manuscript;  
+MZD - study design, manuscript;  
+EL - study design, manuscript;  
+BJK - study design, data collection and analysis, manuscript;
+
+## Disclosures  
+The authors report no relevant disclosures.
+
+## Data Availability  
+Data, analysis scripts, and model code are available at github.com/bjklab.
+
+## Keywords  
+SARS-CoV-2, COVID-19, hospital microbiome
+
+### Acknowledgements:  
+This research, BJK, MJZ, and EL are supported in part by a CDC Cooperative Agreement FOA#CK16-004 "Epicenters for the Prevention of Healthcare Associated Infections”.
+BJK is supported by the National Institute for Allergy and Infectious Diseases (K23 AI121485 and L30 AI120149).
+MJZ is supported by the National Institute for Allergy and Infectious Diseases (K23 AI143925).
+
+### Disclosures:  
+The authors report no conflicts of interest.
+
+### Word Count:    
+?????
+
+-----
+\newpage
+<P style = "page-break-before: always">
+
+## Manuscript
+
+### Abstract
+
+* __Background__:
+
+* __Methods__:
+
+* __Results__:
+
+* __Conclusions__:
+
+ <!-- -----
+\newpage
+<P style = "page-break-before: always">
+
+### Importance
+
+If required by journal. -->
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Introduction:
+
+Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the healthcare and hospital settings.
+In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed uninfected patients developed COVID-19.
+Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].
+
+SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites.
+Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv].
+But the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].
+
+We have previously applied spatial mapping and systematic sampling to define how multidrug-resistant organisms cluster in the hospital environment [site pending ICHE manuscript].
+We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Materials and Methods:
+
+* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA.
+(IRB protocol #8).
+
+* __Causal Model__: We hypothesized that SARS-CoV-2 contamination would decay with physical distance from patient and with time from COVID-19 diagnosis.
+We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects.
+
+* __Specimen Collection and SARS-CoV-2 RT-PCR__:
+
+* __Distance Data Collection__: Distances from the environmental sampling site to the patient bed was recorded using a laser measuring device (Bosch GLM 20).
+Researchers positioned the laser measure at the head of the bed within the room and recorded distances to the in-room ...
+
+* __Definition of Exposures and Outcomes__:
+
+* __Statistical Methods__: Data were organized using R statistical software version 3.6.1 [@R_Core_Team2018-xo], and plots generated using the “ggplot2” package [@Wickham2016-eh].
+We examined how (1) distance from the patient and (2) time since COVID-19 diagnosis relate to the probability of SARS-CoV-2 detection using a binomial model with a logit link.
+We also evaluated mixed-effects model incorporating a random effect for subjects to account for clustering of longitudinal data, and to evaluate inter-subject differences.
+Models were fit using Stan Hamiltonian Monte Carlo (HMC) version 2.21, via the "brms" package with default, weakly-informative priors [@Carpenter2017-ec; @Burkner2017-yr].
+Prior predictive modeling was performed, and models were fit with 4 chains of 1000 iterations, confirmed with HMC diagnostics (no divergent iterations, Rhat statistic < 1.1 for all parameters, and E-BFMI > 0.2) [@McElreath2016-sf; @Gabry2019-le; @Gelman2020-sd].
+We examined parameter distributions at 50%, 80%, and 95% posterior credible intervals to understand the relationship between exposure and outcome variables.
+
+* __Power and Sample Size__: We anticipated that the enrollment of ... subjects would permit ... with type 1 error less than 0.05.
+
+* __Availability of Data__: Data, analysis scripts, and model code are available at https://github.com/bjklab.
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Results:
+
+* __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: ... **Table 1** shows ....
+
+* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We ... **Figure 1** ....
+
+* __SARS-CoV-2 RNA Environmental Contaminations Decays During Prolonged Hospitalization__: We ...
+
+* __SARS-CoV-2 Contamination of the Floor Increases with Pandemic Wave Duration__: We found... **Figure 2** ....
+
+* __SARS-CoV-2 Contamination of High-Touch Objects Depends on Subject Characteristics__: We found... **Table 2** ....
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Discussion:
+
+We sought to define  .
+
+We found  .
+
+The significance  .
+
+Our study has several limitations.  .
+
+In conclusion,   .
+
+-----
+\newpage
+<P style = "page-break-before: always">
+
+### Tables:
+
+#### Table 1:
+
+```{r table1, eval=TRUE, autodep=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
+
+```
+
+__Table 1: Baseline characteristics.__ .
+
+#### Table 2:
+
+```{r table2, eval=TRUE, autodep=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
+
+```
+
+__Table 2: Features that increase risk of high-touch object contamination.__ .
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Figures:
+
+#### Figure 1:
+
+```{r fig1, eval=TRUE, autodep=TRUE, fig.align='center', out.width='90%'}
+knitr::include_graphics(path = "./figs/p_combined_distance_time_unadjusted.png")
+```
+
+__Figure 1: Spatial and temporal effects on SARS-CoV-2 RNA contamination in the healthcare environment.__
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+#### Figure 2:
+
+```{r fig2, eval=TRUE, autodep=TRUE, fig.align='center', out.width='90%'}
+knitr::include_graphics(path = "./figs/p_combined_subject_wave_time_adjusted.png")
+```
+
+__Figure 2: Variation in temporal effects and influence of pandemic duration.__ .
+
+-----  
+\newpage
+<P style = "page-break-before: always">
+
+### Supplemental Figures:
+
+#### Supplemental Figure 1:
+
+```{r sfig1, eval=TRUE, autodep=TRUE, fig.align='center', out.width='60%'}
+#knitr::include_graphics(path = "")
+```
+
+__Supplemental Figure 1: Figure caption__
+
+-----
+\newpage
+<P style = "page-break-before: always">
+
+## References
+
+<div id="refs"></div>
+
+-----
+\newpage
+<P style = "page-break-before: always">
+
+## R Session Information
+
+```{r session_info, eval=TRUE, autodep=TRUE}
+sessionInfo()
+```

### 2021-08-17 manuscript.Rmd
Sketch results description.

@@
 ```{r setup, echo=FALSE, eval=TRUE, include=FALSE, autodep=TRUE}
+
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
@@
 library(gt)
+
+read_csv("./data/dat_long.csv") %>%
+  identity() -> dat_long
+dat_long
+
+dat_long %>%
+  group_by(subject_room_id) %>%
+  filter(subject_room_day == min(subject_room_day)) %>%
+  ungroup() %>%
+  identity() -> dat_first
+dat_first
+
+dat_long %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
+  mutate(scv2_detected = !is.na(copies_max),
+         site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
+                                   grepl("floor",site_descriptor) == FALSE ~ "elevated"),
+         high_touch = touch == "High") %>%
+  #count(site_descriptor, site_category, high_touch)
+  select(subject_room_id, redcap_id, subject_room_day, subject_covid_day, subject_hosp_day, total_study_day, swab_site, unit, site_category, touch, high_touch, distance, copies_max, scv2_detected) %>%
+  mutate(redcap_id = paste0("decon_subject_", redcap_id)) %>%
+  mutate(site_category = stringr::str_to_title(gsub("_"," ",site_category))) %>%
+  distinct() %>%
+  identity() -> dat
+
+m_mvbinom_scv2_distance_fix_category_touch <- read_rds(file = "./models/binomial/m_mvbinom_scv2_distance_fix_category_touch.rds.gz")
+
+m_mvbinom_scv2_time_fix_category_touch <- read_rds(file = "./models/binomial/m_mvbinom_scv2_time_fix_category_touch.rds.gz")
+
+m_mvbinom_scv2_time_category_touch_mix_subject <- read_rds(file = "./models/binomial/m_mvbinom_scv2_time_category_touch_mix_subject.rds.bz2")
+
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave <- read_rds(file = "./models/binomial/m_mvbinom_scv2_time_fix_category_touch_adjust_wave.rds.gz")
+
+tibble(model = list(m_mvbinom_scv2_distance_fix_category_touch, m_mvbinom_scv2_time_fix_category_touch, m_mvbinom_scv2_time_category_touch_mix_subject, m_mvbinom_scv2_time_fix_category_touch_adjust_wave), model_name = c("Distance Fixed", "Time Fixed", "Time Subject Random", "Time Pandemic Wave Adjusted")) %>%
+  mutate(model_summary = map(model, .f = ~ brms::posterior_summary(.x) %>% as_tibble(rownames = "param"))) %>%
+  select(model_name, model_summary) %>%
+  unnest(cols = c(model_summary)) %>%
+  mutate_if(.predicate = ~ is.numeric(.x), .funs = list("exp" = ~ exp(.x))) %>%
+  identity() -> m_combined_summary
+
 ```
@@
 
-* __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: ... **Table 1** shows ....
+* __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: We collected `r nrow(dat)` specimens from `r nrow(count(dat,subject_room_id))` unique subject-room pairs, comprising `r nrow(count(dat,redcap_id))` unique subjects (i.e., `r nrow(filter(count(distinct(select(dat,subject_room_id,redcap_id)),redcap_id),n>1))` subjects were transferred at least once during the observation period).
+The median number of sampling events from each subject-room pair was `r round(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r round(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`).
+**Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, and proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-PCR....
 
-* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We ... **Figure 1** ....
+* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces.
+There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [site ICEMAN ICHE paper], but the suface type was significantly associated with the probability of detection.
+Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces.
+High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces.
+**Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
 
-* __SARS-CoV-2 RNA Environmental Contaminations Decays During Prolonged Hospitalization__: We ...
+* __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We ...
 
@@
 
+dat %>%
+  left_join(distinct(select(dat_long, swab_site, site_descriptor)), by = "swab_site") %>%
+  mutate(high_touch = case_when(high_touch == TRUE ~ "High Touch",
+                                high_touch == FALSE ~ "Low Touch")) %>%
+  mutate(site_descriptor = stringr::str_to_title(string = gsub("_"," ",gsub("1|2","",site_descriptor)))) %>%
+  group_by(site_category, site_descriptor, high_touch) %>%
+  summarise(unique_subjects = n_distinct(subject_room_id),
+            unique_specimens = n(),
+            prop_positive = sum(!is.na(copies_max)) / n()) %>%
+  ungroup() %>%
+  arrange(desc(prop_positive)) %>%
+  gt::gt() %>%
+  gt::cols_label(site_category = "Height", site_descriptor = "Location", high_touch = html("Contact"), prop_positive = html("SARS-CoV-2<br>Detected"), unique_subjects = html("Subjects<br>Sampled"), unique_specimens = html("Surfaces<br>Sampled")) %>%
+  gt::fmt_percent(columns = prop_positive, decimals = 1) %>%
+  identity() -> dat_gt
+dat_gt
+
+dat_gt %>%
+  gt::as_raw_html() %>%
+  write_lines("./tabs/surfaces_sampled.html")
+
 ```
 
-__Table 1: Baseline characteristics.__ .
+__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ .
 
@@
 
-__Table 2: Features that increase risk of high-touch object contamination.__ .
+__Table 2: Subject characteristics that increase risk of high-touch object contamination.__ .
 

### 2021-08-17 manuscript.Rmd
Draft results and figure legends.

@@
 library(tidybayes)
+library(glue)
 library(gt)
@@
 The median number of sampling events from each subject-room pair was `r round(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r round(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`).
-**Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, and proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-PCR....
+**Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, the proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-qPCR, and the SARS-CoV-2 copy numbers measured.
+We did not observe significant differences in copy numbers across the surface types, but the proportion with detectable SARS-CoV-2 did vary widely.
 
@@
 
-* __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We ...
+* __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We evaluated the effect of time from COVID-19 diagnosis on the detection of SARS-CoV-2 RNA in the hospital environment via a multivariable logistic regression model relating time from COVID-19 diagnosis, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
+We found that the probability of detection decreased as time from COVID-19 diagnosis increased, with OR `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q97.5_exp) %>% round(.,2)`) per day since COVID-19 diagnosis.
+The most prominent effects were observed on high-touch elevated surfaces and floor surfaces (**Figure 1B**).
 
-* __SARS-CoV-2 Contamination of the Floor Increases with Pandemic Wave Duration__: We found... **Figure 2** ....
+* __SARS-CoV-2 Contamination of the Floor Increases with Pandemic Wave Duration__: To understand how subjects differed in the effect of time from COVID-19 diagnosis on SARS-CoV-2 contamination of the hospital environment, we re-fit the above model with the addition of random intercepts and slopes to account for subject-level differences.
+**Figure 2A** depicts the best estimate of the effect of time from COVID-19 diagnosis on the contamination in each subject's hospital room.
+In the figure, the lines representing each subject are colored by when they were enrolled during the course of the local, second-wave COVID-19 case surge.
+We observed that the contamination of high-touch elevated surfaces did not appear to relate to time from the start of the local case surge.
+However, contamination of floor surfaces appeared to be greater among subjects enrolled later in the course of the local case surge.
+To investigate this further, we added an interaction term to the model and found that each day since the start of the start of the local second wave increased the contamination of floor surfaces with OR `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q97.5_exp) %>% round(.,2)`) (**Figure 2B**).
 
@@
             unique_specimens = n(),
-            prop_positive = sum(!is.na(copies_max)) / n()) %>%
+            prop_positive = sum(!is.na(copies_max)) / n(),
+            med_copies = median(log10(copies_max), na.rm = TRUE),
+            q25_copies = quantile(log10(copies_max), probs = 0.25, na.rm = TRUE),
+            q75_copies = median(log10(copies_max),  probs = 0.75, na.rm = TRUE),
+            copies_summary = glue("{round(med_copies,1)} ({round(q25_copies,1)} - {round(q75_copies,1)})")) %>%
   ungroup() %>%
   arrange(desc(prop_positive)) %>%
+  select(- contains("_copies")) %>%
   gt::gt() %>%
-  gt::cols_label(site_category = "Height", site_descriptor = "Location", high_touch = html("Contact"), prop_positive = html("SARS-CoV-2<br>Detected"), unique_subjects = html("Subjects<br>Sampled"), unique_specimens = html("Surfaces<br>Sampled")) %>%
+  gt::cols_label(site_category = "Height", site_descriptor = "Location", high_touch = html("Contact"), unique_subjects = html("Subjects<br>Sampled"), unique_specimens = html("Surfaces<br>Sampled"), prop_positive = html("SARS-CoV-2<br>Detected"), copies_summary = html("Median (IQR) log<sub>10</sub><br>Copies by RT-qPCR")) %>%
   gt::fmt_percent(columns = prop_positive, decimals = 1) %>%
@@
 
-__Figure 1: Spatial and temporal effects on SARS-CoV-2 RNA contamination in the healthcare environment.__
+__Figure 1: Spatial and temporal effects on SARS-CoV-2 RNA contamination in the healthcare environment.__ (A) Distance from the head of the patient's bed is shown on the horizontal axis.
+The vertical axis depicts the probability of SARS-CoV-2 detection by RT-qPCR according to a logistic regression model incorporating the surface elevation and touch.
+(B) Days from diagnosis with COVID-19 are shown on the horizontal axis.
+The vertical axis depicts the probability of SARS-CoV-2 detection by RT-qPCR according to a logistic regression model incorporating the surface elevation and touch.
+For both plots, the black line shows the best estimate, and shading indicates 50%, 80%, and 95% posterior credible intervals.
 
@@
 
-__Figure 2: Variation in temporal effects and influence of pandemic duration.__ .
+__Figure 2: Variation in temporal effects and influence of pandemic duration.__ (A) Days from diagnosis with COVID-19 are shown on the horizontal axis.
+The vertical axis depicts the probability of SARS-CoV-2 detection by RT-qPCR according to a logistic regression model incorporating the site elevation and touch, with random subject-level effects.
+Each line represents the best estimate for a single subjects; lines are colored according to the number of days from the start of the local COVID-19 second case wave.
+(B) Days from the local start of the second COVID-19 wave are shown on the horizontal axis.
+The vertical axis depicts the probability of SARS-CoV-2 detection by RT-qPCR according to a logistic regression model incorporating the days from COVID-19 diagnosis, surface elevation, and touch, with an interaction term between the days from the local start of the second COVID-19 wave and the surface type.
+The black line shows the best estimate, and shading indicates 50%, 80%, and 95% posterior credible intervals.
 

### 2021-08-17 scripts/model_manuscript.R

@@ -420,7 +420,7 @@ m_mvbinom_scv2_time_category_touch_mix_subject_fitted %>%
   colorspace::scale_color_continuous_sequential(palette = "YlGnBu") +
   labs(x = "Days after COVID-19 Diagnosis",
        y = "Probability of SARS-CoV-2<br>Detection by RT-PCR",
-       color = "Days from<br>Start of<br>Enrollment") +
+       color = "Days from<br>Start of<br>Local<br>COVID-19<br>Wave") +
   theme_bw() +
   theme(strip.text = ggtext::element_markdown(color = "black", size = 10),
         axis.text.x = ggtext::element_markdown(color = "black"),
@@ -556,7 +556,7 @@ m_mvbinom_scv2_time_fix_category_touch_adjust_wave_fitted %>%
   stat_lineribbon() +
   facet_wrap(facets = ~ site_category + high_touch, scales = "fixed") +
   #facet_wrap(facets = ~ site_category) +
-  scale_fill_brewer(palette = "Purples") +
+  scale_fill_brewer(palette = "Blues") +
   scale_y_continuous(limits = c(0,1)) +
   labs(x = "Days from Start of Second COVID-19 Wave",
        y = "Probability of SARS-CoV-2<br>Detection by RT-PCR",
@@ -592,7 +592,7 @@ p_scv2_mvbinomial_subject_covid_day_site_category_touch_adjust_wave
 
 #' patchwork plot joining fixed-effects and random-subject-effects models
 
-((p_scv2_mvbinomial_distance_site_category_touch_mix_subject_lines) / 
+((p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines) / 
     (p_scv2_mvbinomial_subject_covid_day_site_category_touch_adjust_wave)) + #theme(strip.text = element_blank()))) +
   #patchwork::plot_layout(guides = 'collect') +
   patchwork::plot_annotation(tag_levels = "A") -> p_combined_subject_wave_time_adjusted

### 2021-09-06 manuscript.Rmd
Manuscript Additions

@@
 * __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA.
-(IRB protocol #8).
+(IRB protocol #843273).
+Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability.
+All rooms underwent daily cleaning, according to hospital protocol.
+Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects (one wipe per surface).
+Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).
 
@@
 
-* __Specimen Collection and SARS-CoV-2 RT-PCR__:
+* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital (i.e., hospital days 1,2,7,14,21,and 28).
+Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
+Multiple surfaces were swabbed within each patient room using a Copan FLOQ(TM) swab and a sterile template (20cm2).
+Specific surfaces are listed in Table 1.
+Swabs were immediately placed into buffered AVL and frozen prior to further processing.
+Specimens then underwent RNA extraction and RT-PCR targeting the SARS-CoV-2 N1 region in triplicate.
 
-* __Distance Data Collection__: Distances from the environmental sampling site to the patient bed was recorded using a laser measuring device (Bosch GLM 20).
-Researchers positioned the laser measure at the head of the bed within the room and recorded distances to the in-room ...
+* __Distance Data Collection__: Distance from the environmental sampling site to the patient's head of bed was recorded using a laser measuring device (Bosch GLM 20).
+Templates were affixed to each surface and the linear distance from the patient's head of bed to the sample site was measured.
 
 * __Definition of Exposures and Outcomes__: 
+Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
+The primary study outcome of interest, detection of SARS-CoV-2 by PCR assay, was determined by detection of any quantifiable RNA.
 
@@
 
-* __Power and Sample Size__: We anticipated that the enrollment of ... subjects would permit ... with type 1 error less than 0.05.
+* __Power and Sample Size__: We anticipated that the enrollment of a minumum of 100 subjects would permit detection of a minimum of a 10% decrease probability of surface detection between distance groups with type 1 error less than 0.05.
 
@@
 
-We sought to define  .
+We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of surfaces within COVID-19 patient rooms.
+We found that distance from the patient's head of bed did not have a significant impact on the probability of detection of SARS-CoV-2 RNA.
+We had hypothesized that the probability of RNA detection with decrease with distance.
+However, we found that detection did not vary significantly among any surface group with increased distance.
+This is likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch.
+We observed large differences in the probability of detecting surface contamination by surface type.
+Floors had high probability of contamination, which did not vary throughout the room.
+This is likely related to both the role of patient and HCP feet as well as the physical action of mopping which spread body fluids containing viral RNA within the patient's room.
+In contrast, walls had significantly lower probability of contamination.
+This may be due to the decreased settling of respiratory droplets on vertical surfaces.
+This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020].
 
-We found  .
+While we did not observe a measurable impact of distance on the probability of surface contamination with SARS-CoV-2 RNA, we did detect a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis.
+A similar finding was described by Chia et al [@Chia2020], where patients over 7 days from illness had lower probability of surface contamination within rooms.
+Interestingly, while the probability of floor contamination decreased with time from diagnosis, the probability remained relatively high at later time points.
+Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a significant proportion of floor contamination.
+Viral RNA carried on the feet of healthcare personnel or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased.
+Guo et al [@Guo2020] described that floor contamination was discovered in areas where clinical care was not performed, and suggested that it was spread by the feet of HCP.
 
-The significance  .
+Our findings are significant as they reveal that varied distances within a patient room have similar probability of surface contamination.
+This finding supports CDC recommended infection control practices [@CDCInfectionControl].
+Additionally, our findings highlight the role of touch in surface contamination of patient rooms and the likelihood of extrinsic contamination on SARS-CoV-2 units.
 
-Our study has several limitations.  .
+Our study has several limitations.
+First, our study used quantitative PCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
+Secondly, we did not control for cleaning methods on study units.
+However, cleaning practices did not undergo significant changes during the study time period and sampling was performed routinely prior to daily cleaning.
+Finally, while we explored patient characteristics impacted the decay of surface contamination with time and distance, there was limited variability of some measures, making independent associations with viral detection less feasible.
 

### 2021-09-14 scripts/model_manuscript.R
Sketch subject-level feature model.

@@ -558,7 +558,7 @@ m_mvbinom_scv2_time_fix_category_touch_adjust_wave_fitted %>%
   #facet_wrap(facets = ~ site_category) +
   scale_fill_brewer(palette = "Blues") +
   scale_y_continuous(limits = c(0,1)) +
-  labs(x = "Days from Start of Second COVID-19 Wave",
+  labs(x = "Days from Start of Local COVID-19 Wave",
        y = "Probability of SARS-CoV-2<br>Detection by RT-PCR",
        fill = "Posterior<br>Credible<br>Interval") +
   theme_bw() +
@@ -610,8 +610,111 @@ p_combined_subject_wave_time_adjusted
 
 
 
+#' ####################################################
+#' ####################################################
+#' EXPLORE FOR SUBJECT-LEVEL FEATURES a/w CONTAMINATION
+#' ####################################################
+#' ####################################################
+
+library(projpred)
+dat_subject <- read_csv("./data/dat_subjects.csv")
+dat_subject
+
+dat_subject %>%
+  summarise_at(.vars = vars(-redcap_id), .funs = list(count = ~ sum(.x), proportion = ~ sum(.x)/length(.x))) %>%
+  pivot_longer(cols = everything()) %>%
+  mutate(measure = stringr::str_extract(string = name, pattern = "count|proportion"),
+         name = gsub("_count|_proportion","",name)) %>%
+  pivot_wider(id_cols = "name", names_from = "measure", values_from = "value") %>%
+  gt() %>%
+  gt::fmt_percent(columns = proportion)
 
 
+dat_subject %>%
+  count(mech_vent_ever, supp_oxy_ever, bipap_cpap_ever, steroid, remdesivir, name = "count") %>%
+  gt() %>%
+  gt::opt_all_caps() %>%
+  gtExtras::gt_color_rows(columns = -count, palette = "basetheme::brutal", direction = 1, use_paletteer = TRUE) %>%
+  gtExtras::gt_theme_538() %>%
+  identity()
+  #gt::gtsave(./tabs/decontam_check_subject_data.html")
+
+
+dat_long %>%
+  # exclude bathroom sites and nursing sites (different distance scheme)
+  filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
+  mutate(scv2_detected = !is.na(copies_max),
+         site_category = case_when(grepl("floor",site_descriptor) == TRUE ~ "floor",
[…]

### 2021-09-14 scripts/model_manuscript.R
Model impact of COVID-19 severity.

@@ -612,32 +612,43 @@ p_combined_subject_wave_time_adjusted
 
 #' ####################################################
 #' ####################################################
-#' EXPLORE FOR SUBJECT-LEVEL FEATURES a/w CONTAMINATION
+#' COVID SEVERITY IMPACT on ENVIRONMENTAL CONTAMINATION
 #' ####################################################
 #' ####################################################
 
-library(projpred)
-dat_subject <- read_csv("./data/dat_subjects.csv")
+dat_subject <- read_csv("./data/dat_subjects.csv") %>%
+  mutate(covid_severity = case_when(mech_vent_ever == FALSE & high_flow_nc_ever  == FALSE & bipap_cpap_ever == FALSE & supp_oxy_ever == FALSE ~ "No Oxygen Support",
+                                    mech_vent_ever == FALSE & high_flow_nc_ever  == FALSE & bipap_cpap_ever == FALSE & supp_oxy_ever == TRUE ~ "Only Nasal or Facemask Oxygen Support",
+                                    mech_vent_ever == TRUE | high_flow_nc_ever  == TRUE | bipap_cpap_ever == TRUE ~ "Significant Oxygen Support"),
+         covid_severity = factor(covid_severity, levels = c("No Oxygen Support", "Only Nasal or Facemask Oxygen Support", "Significant Oxygen Support")))
 dat_subject
 
+
 dat_subject %>%
-  summarise_at(.vars = vars(-redcap_id), .funs = list(count = ~ sum(.x), proportion = ~ sum(.x)/length(.x))) %>%
+  summarise_at(.vars = vars(-redcap_id, -covid_severity), .funs = list(count = ~ sum(.x), proportion = ~ sum(.x)/length(.x))) %>%
   pivot_longer(cols = everything()) %>%
   mutate(measure = stringr::str_extract(string = name, pattern = "count|proportion"),
          name = gsub("_count|_proportion","",name)) %>%
   pivot_wider(id_cols = "name", names_from = "measure", values_from = "value") %>%
+  arrange(desc(proportion)) %>%
   gt() %>%
-  gt::fmt_percent(columns = proportion)
+  gt::fmt_percent(columns = proportion) %>%
+  gtExtras::gt_color_rows(columns = -name, palette = viridis::turbo(n = 10), use_paletteer = FALSE) %>%
+  gtExtras::gt_theme_538() %>%
+  identity()
[…]

### 2021-09-14 manuscript.Rmd
Add COVID-19 severity model to results.

@@
 
+dat_subject <- read_csv("./data/dat_subjects.csv") %>%
+  mutate(covid_severity = case_when(mech_vent_ever == FALSE & high_flow_nc_ever  == FALSE & bipap_cpap_ever == FALSE & supp_oxy_ever == FALSE ~ "No Oxygen Support",
+                                    mech_vent_ever == FALSE & high_flow_nc_ever  == FALSE & bipap_cpap_ever == FALSE & supp_oxy_ever == TRUE ~ "Only Nasal or Facemask Oxygen Support",
+                                    mech_vent_ever == TRUE | high_flow_nc_ever  == TRUE | bipap_cpap_ever == TRUE ~ "Significant Oxygen Support"),
+         covid_severity = factor(covid_severity, levels = c("No Oxygen Support", "Only Nasal or Facemask Oxygen Support", "Significant Oxygen Support")))
+dat_subject
+
 m_mvbinom_scv2_distance_fix_category_touch <- read_rds(file = "./models/binomial/m_mvbinom_scv2_distance_fix_category_touch.rds.gz")
@@
 
-tibble(model = list(m_mvbinom_scv2_distance_fix_category_touch, m_mvbinom_scv2_time_fix_category_touch, m_mvbinom_scv2_time_category_touch_mix_subject, m_mvbinom_scv2_time_fix_category_touch_adjust_wave), model_name = c("Distance Fixed", "Time Fixed", "Time Subject Random", "Time Pandemic Wave Adjusted")) %>%
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity <- read_rds(file = "./models/binomial/m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity.rds.gz")
+
+tibble(model = list(m_mvbinom_scv2_distance_fix_category_touch, m_mvbinom_scv2_time_fix_category_touch, m_mvbinom_scv2_time_category_touch_mix_subject, m_mvbinom_scv2_time_fix_category_touch_adjust_wave, m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity), model_name = c("Distance Fixed", "Time Fixed", "Time Subject Random", "Time Pandemic Wave Adjusted", "COVID-19 Severity")) %>%
   mutate(model_summary = map(model, .f = ~ brms::posterior_summary(.x) %>% as_tibble(rownames = "param"))) %>%
@@
 
-* __SARS-CoV-2 Contamination of High-Touch Objects Depends on Subject Characteristics__: We found... **Table 2** ....
+* __SARS-CoV-2 Contamination of Increases with COVID-19 Disease Severity__: To understand how COVID-19 disease severity could impact environmental contamination with SARS-CoV-2 RNA, we categorized subjects based on their required oxygen support: `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[1]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[1]`) subjects who required no oxygen support, `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[2]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[2]`) subjects who required only nasal or facemask oxygen support, and `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[3]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[3]`) subjects who required significant (i.e., high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation) oxygen support.
+Adjusting for the site, time from COVID-19 diagnosis, and time from start of the local COVID-19 wave, we found that the use of significant oxygen support was associated with increased environmental SARS-CoV-2 RNA contamination, with OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% round(.,2)` ((95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% round(.,2)`)) relative to the no-oxygen-support reference group.
+**Figure 3** shows the predicted probabilities of SARS-CoV-2 environmental contamination as it relates to COVID-19 severity across all measured surfaces, adjusted for time from COVID-19 diagnosis and time from the start of the local COVID-19 wave.
 
@@
 
-#### Table 2:
-
-```{r table2, eval=TRUE, autodep=TRUE, error=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
-
-```
-
-__Table 2: Subject characteristics that increase risk of high-touch object contamination.__ .
-
 -----  
@@
 
-### Supplemental Figures:
+#### Figure 3:
 
-#### Supplemental Figure 1:
-
-```{r sfig1, eval=TRUE, autodep=TRUE, fig.align='center', out.width='60%'}
-#knitr::include_graphics(path = "")
+```{r fig3, eval=TRUE, autodep=TRUE, fig.align='center', out.width='50%'}
+knitr::include_graphics(path = "./figs/p_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity.png")
 ```
 
-__Supplemental Figure 1: Figure caption__
+__Figure 3: Impact of COVID-19 disease severity on environmental contamination.__ To evaluate the effect of COVID-19 disease severity on environmental contamination, we compared the probability of SARS-CoV-2 RNA detection by RT-PCR (horizontal axis) across subjects with three levels of necessary oxygen support (vertical axis) at each sampling site.
+The point shows the best estimate and the sigment indicates the 95% posterior credible interval, with adjustment for time from COVID-19 diagnosis and time from the start of the local COVID-19 wave.
 

### 2021-09-14 scripts/model_manuscript.R

@@ -814,6 +814,96 @@ p_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity
 
 
 
+
+
+
+
+
+#' ####################################################
+#' ####################################################
+#' WHAT'S THE IMPACT OF INTERACTION?
+#' ####################################################
+#' ####################################################
+library(loo)
+
+#' run binomial model
+dat %>%
+  select(scv2_detected, subject_covid_day, total_study_day, site_category, high_touch, covid_severity) %>%
+  brm(formula = scv2_detected ~ 1 + subject_covid_day + site_category * total_study_day + high_touch + site_category * covid_severity,
+      data = .,
+      family = bernoulli,
+      #prior = prior(horseshoe(scale_global = 0.2, scale_slab = 1), class=b),
+      chains = 4,
+      cores = 4,
+      control = list("adapt_delta" = 0.999, max_treedepth = 10),
+      backend = "cmdstanr",
+      seed = 16) -> m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X
+
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X %>% write_rds(file = "./models/binomial/m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X.rds.gz", compress = "gz")
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X <- read_rds(file = "./models/binomial/m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X.rds.gz")
+
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X
+rstan::check_hmc_diagnostics(m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X$fit)
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X %>% pp_check()
+
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X %>%
+  posterior_summary() %>%
+  as_tibble(rownames = "param") %>%
+  gt::gt() %>%
+  gt::fmt_number(columns = 2:5, n_sigfig = 3)
+
+
+#' fitted
+m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity_X$data %>%
+  as_tibble() %>%
+  expand(subject_covid_day = modelr::seq_range(subject_covid_day, n = 10),
[…]

### 2021-09-16 manuscript.Rmd
Manuscript and Bib File Additions

@@
 
-* __Background__:
+* __Background__: SARS-CoV-2 RNA has been described to contaminate surfaces within healthcare settings.
+However, little is known regarding the role of distance and time in surface contamination in a clinical setting.
 
-* __Methods__:
+* __Methods__: We performed longitudinal sampling of surfaces within COVID-19 patient rooms while measuring the distance from each surface to the patient bed.
+We performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA and modeled the role of distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.
 
-* __Results__:
+* __Results__: Probability of detecting SARS-CoV-2 RNA did not vary with distance.
+However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: odds ratio [OR] 67.77 (95% confidence interval [CI] 36.34 to 131.47, high-touch elevated: OR 7.39 (95% CI 4.31 to 13.09)).
+Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory support had greater odds of  surface contamination (OR 1.76 (95% CrI 1.2 to 2.57)).
+While the probability of surface contamination decayed with prolonged hospitalization, probability of floor detection increased with time over the pandemic.
 
-* __Conclusions__:
+* __Conclusions__: SARS-CoV-2 surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time predicted surface deposition.
 
@@
 Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv].
-But the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].
+However, the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].
 
@@
 All rooms underwent daily cleaning, according to hospital protocol.
-Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects (one wipe per surface).
+Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface.
 Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).
@@
 Multiple surfaces were swabbed within each patient room using a Copan FLOQ(TM) swab and a sterile template (20cm2).
-Specific surfaces are listed in Table 1.
+Specific surfaces are listed in Table 1 with high-touch surfaces selected using structured observations prior to study start.
 Swabs were immediately placed into buffered AVL and frozen prior to further processing.
@@
 * __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces.
-There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [site ICEMAN ICHE paper], but the suface type was significantly associated with the probability of detection.
+There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection.
 Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces.
@@
 
-We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of surfaces within COVID-19 patient rooms.
-We found that distance from the patient's head of bed did not have a significant impact on the probability of detection of SARS-CoV-2 RNA.
-We had hypothesized that the probability of RNA detection with decrease with distance.
-However, we found that detection did not vary significantly among any surface group with increased distance.
-This is likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch.
+****ADD Discussion of our earlier ICHE paper
+
+We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of  within COVID-19 patient rooms.
+We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA.
+We had hypothesized that the probability of RNA detection would decrease with distance.
+This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@KElly2021].
+However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied.
+This is most likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch.
 We observed large differences in the probability of detecting surface contamination by surface type.
 Floors had high probability of contamination, which did not vary throughout the room.
-This is likely related to both the role of patient and HCP feet as well as the physical action of mopping which spread body fluids containing viral RNA within the patient's room.
-In contrast, walls had significantly lower probability of contamination.
-This may be due to the decreased settling of respiratory droplets on vertical surfaces.
+This is likely related to both the role of patient and healthcare personnel (HCP) foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room.
 This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020].
+In contrast to floors, walls had significantly lower probability of contamination.
+This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
 
@@
 Interestingly, while the probability of floor contamination decreased with time from diagnosis, the probability remained relatively high at later time points.
-Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a significant proportion of floor contamination.
-Viral RNA carried on the feet of healthcare personnel or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased.
-Guo et al [@Guo2020] described that floor contamination was discovered in areas where clinical care was not performed, and suggested that it was spread by the feet of HCP.
+Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a proportion of floor contamination.
+Viral RNA carried on the feet of HCP or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased.
+This is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020].
+Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
 
-Our findings are significant as they reveal that varied distances within a patient room have similar probability of surface contamination.
-This finding supports CDC recommended infection control practices [@CDCInfectionControl].
-Additionally, our findings highlight the role of touch in surface contamination of patient rooms and the likelihood of extrinsic contamination on SARS-CoV-2 units.
+With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms.
+This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aeresol production associated with respiratory procedures.
+In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination.
+Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
 
@@
 Secondly, we did not control for cleaning methods on study units.
-However, cleaning practices did not undergo significant changes during the study time period and sampling was performed routinely prior to daily cleaning.
-Finally, while we explored patient characteristics impacted the decay of surface contamination with time and distance, there was limited variability of some measures, making independent associations with viral detection less feasible.
+However, cleaning practices did not undergo systematic changes during the study time period and sampling was performed routinely prior to daily cleaning.
+Finally, we did not perform matching sampling of occupying patients to explore the role of patient viral load and surface contamination.
 
-In conclusion,   .
+In conclusion, SARS-CoV-2 RNA surface deposition did not vary with distance within patient rooms.
+However, surface type, severity of illness, and time predicted detection of surface contamination.
+Future studies are needed to determine the role of surface contamination in risk of infection among patients and HCP.
 

### 2021-09-16 manuscript.Rmd
Quick Manuscript Correction

@@
 
-****ADD Discussion of our earlier ICHE paper
-
 We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of  within COVID-19 patient rooms.
@@
 We had hypothesized that the probability of RNA detection would decrease with distance.
-This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@KElly2021].
+This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021].
 However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied.

### 2021-09-17 manuscript.Rmd
Re-run COVID severity model, correct to first only.

@@
 -----
+
 \newpage
-<P style = "page-break-before: always">
 
@@
 
-* __Background__: SARS-CoV-2 RNA has been described to contaminate surfaces within healthcare settings.
-However, little is known regarding the role of distance and time in surface contamination in a clinical setting.
+* __Background__: SARS-CoV-2 is known to contaminate surfaces within healthcare settings, but the spatial and temporal extent of environmental contamination has not been precisely defined.
+We sought to define SARS-CoV-2 contamination of different surface types and how contamination changes over time.
 
@@
 
-* __Results__: Probability of detecting SARS-CoV-2 RNA did not vary with distance.
-However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: odds ratio [OR] 67.77 (95% confidence interval [CI] 36.34 to 131.47, high-touch elevated: OR 7.39 (95% CI 4.31 to 13.09)).
-Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory support had greater odds of  surface contamination (OR 1.76 (95% CrI 1.2 to 2.57)).
-While the probability of surface contamination decayed with prolonged hospitalization, probability of floor detection increased with time over the pandemic.
+* __Results__: The probability of detecting SARS-CoV-2 RNA did not vary with distance.
+However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated: OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+The probability of surface contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
 
@@
 \newpage
-<P style = "page-break-before: always">
 
@@
 
-We have previously applied spatial mapping and systematic sampling to define how multidrug-resistant organisms cluster in the hospital environment [site pending ICHE manuscript].
+We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021-vm].
 We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.
@@
 
-* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA.
-(IRB protocol #843273).
+* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273).
 Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability.
@@
 * __Causal Model__: We hypothesized that SARS-CoV-2 contamination would decay with physical distance from patient and with time from COVID-19 diagnosis.
-We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects.
+We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects, and that increased COVID-19 disease severity would be associated with greater environmental contamination.
 
@@
 * __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: We collected `r nrow(dat)` specimens from `r nrow(count(dat,subject_room_id))` unique subject-room pairs, comprising `r nrow(count(dat,redcap_id))` unique subjects (i.e., `r nrow(filter(count(distinct(select(dat,subject_room_id,redcap_id)),redcap_id),n>1))` subjects were transferred at least once during the observation period).
-The median number of sampling events from each subject-room pair was `r round(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r round(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`).
+The median number of sampling events from each subject-room pair was `r signif(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r signif(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`).
 **Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, the proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-qPCR, and the SARS-CoV-2 copy numbers measured.
@@
 * __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces.
-There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection.
-Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces.
-High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces.
+There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021-vm], but the suface type was significantly associated with the probability of detection.
+Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces.
+High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces.
 **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
@@
 * __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We evaluated the effect of time from COVID-19 diagnosis on the detection of SARS-CoV-2 RNA in the hospital environment via a multivariable logistic regression model relating time from COVID-19 diagnosis, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
-We found that the probability of detection decreased as time from COVID-19 diagnosis increased, with OR `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q97.5_exp) %>% round(.,2)`) per day since COVID-19 diagnosis.
+We found that the probability of detection decreased as time from COVID-19 diagnosis increased, with OR `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q97.5_exp) %>% signif(.,3)`) per day since COVID-19 diagnosis.
 The most prominent effects were observed on high-touch elevated surfaces and floor surfaces (**Figure 1B**).
@@
 However, contamination of floor surfaces appeared to be greater among subjects enrolled later in the course of the local case surge.
-To investigate this further, we added an interaction term to the model and found that each day since the start of the start of the local second wave increased the contamination of floor surfaces with OR `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q97.5_exp) %>% round(.,2)`) (**Figure 2B**).
+To investigate this further, we added an interaction term to the model and found that each day since the start of the start of the local second wave increased the contamination of floor surfaces with OR `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q97.5_exp) %>% signif(.,3)`) (**Figure 2B**).
 
-* __SARS-CoV-2 Contamination of Increases with COVID-19 Disease Severity__: To understand how COVID-19 disease severity could impact environmental contamination with SARS-CoV-2 RNA, we categorized subjects based on their required oxygen support: `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[1]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[1]`) subjects who required no oxygen support, `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[2]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[2]`) subjects who required only nasal or facemask oxygen support, and `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[3]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[3]`) subjects who required significant (i.e., high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation) oxygen support.
-Adjusting for the site, time from COVID-19 diagnosis, and time from start of the local COVID-19 wave, we found that the use of significant oxygen support was associated with increased environmental SARS-CoV-2 RNA contamination, with OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% round(.,2)` ((95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% round(.,2)`)) relative to the no-oxygen-support reference group.
+* __SARS-CoV-2 Contamination of Increases with COVID-19 Disease Severity__: To understand how COVID-19 disease severity could impact environmental contamination with SARS-CoV-2 RNA, we categorized subjects based on their required oxygen support: `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[1]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[1]`) subjects who required no oxygen support, `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[2]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[2]`) subjects who required only nasal or facemask oxygen support, and `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[3]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[3]`) subjects who required significant (i.e., high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation) oxygen support.
+Adjusting for the site, time from COVID-19 diagnosis, and time from start of the local COVID-19 wave, we found that the use of significant oxygen support was associated with increased environmental SARS-CoV-2 RNA contamination, with OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`) relative to the no-oxygen-support reference group.
 **Figure 3** shows the predicted probabilities of SARS-CoV-2 environmental contamination as it relates to COVID-19 severity across all measured surfaces, adjusted for time from COVID-19 diagnosis and time from the start of the local COVID-19 wave.
@@
 
-We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of  within COVID-19 patient rooms.
+We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms.
 We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA.
 We had hypothesized that the probability of RNA detection would decrease with distance.
-This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021].
+This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm].
 However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied.
@@
 With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms.
-This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aeresol production associated with respiratory procedures.
+This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures.
 In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination.
-Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
+Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].
 
@@
 
-__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ .
+__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are reproted.
 
 -----
+
 \newpage
-<P style = "page-break-before: always">
 

### 2021-09-17 scripts/model_manuscript.R

@@ -651,7 +651,7 @@ dat_subject %>%
   #gt::gtsave("./tabs/decontam_check_subject_data.html")
 
 
-dat_long %>%
+dat_first %>%
   # exclude bathroom sites and nursing sites (different distance scheme)
   filter(grepl("bathroom|sink|toilet|nurse",site_descriptor) == FALSE) %>%
   mutate(scv2_detected = !is.na(copies_max),
@@ -702,18 +702,6 @@ m_mvbinom_scv2_time_category_touch_mix_subject_fitted %>%
 p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines_severity
 
 
-# p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines %>%
-#   ggsave(filename = "./figs/p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines.pdf", height = 5, width = 8, units = "in")
-# p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines %>%
-#   ggsave(filename = "./figs/p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines.png", height = 5, width = 8, units = "in", dpi = 600)
-# p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines %>%
-#   ggsave(filename = "./figs/p_scv2_mvbinomial_time_site_category_touch_mix_subject_lines.svg", height = 5, width = 8, units = "in")
-
-
-
-
-
-
 
 #' ####################################################
 #' generative model for SARS-CoV-2 contamination ~ days from COVID diagnosis, adjusted for total days since local second wave
@@ -732,19 +720,19 @@ dat %>%
 
 
 #' run binomial model
-dat %>%
-  select(scv2_detected, subject_covid_day, total_study_day, site_category, high_touch, covid_severity) %>%
-  brm(formula = scv2_detected ~ 1 + subject_covid_day + site_category * total_study_day + high_touch + covid_severity,
-      data = .,
-      family = bernoulli,
-      #prior = prior(horseshoe(scale_global = 0.2, scale_slab = 1), class=b),
-      chains = 4,
-      cores = 4,
-      control = list("adapt_delta" = 0.999, max_treedepth = 10),
-      backend = "cmdstanr",
-      seed = 16) -> m_mvbinom_scv2_time_fix_category_touch_adjust_wave_covid_severity
-
[…]

### 2021-09-17 manuscript.Rmd
Expand manuscript discussion.

@@
 * __Results__: The probability of detecting SARS-CoV-2 RNA did not vary with distance.
-However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated: OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-The probability of surface contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
+However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
 
-* __Conclusions__: SARS-CoV-2 surface deposition did not vary with distance within patient rooms. However, surface type, severity of illness, and time predicted surface deposition.
+* __Conclusions__: Distance from patient's bed did not predict SARS-CoV-2 RNA deposition in patient rooms, but surface type, severity of illness, and time predicted surface deposition.
 
@@
 Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv].
-However, the spatial extent of contamination -- and how contamination changes during the course of hospitalization -- remain less well understood [@Mody2021-ag].
+However, the spatial extent of contamination and how contamination changes during the course of hospitalization remain less well understood [@Mody2021-ag].
 
@@
 Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface.
-Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).
+Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.
+Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
 
-* __Causal Model__: We hypothesized that SARS-CoV-2 contamination would decay with physical distance from patient and with time from COVID-19 diagnosis.
+* __Causal Model__: We hypothesized that SARS-CoV-2 contamination of the healthcare environment would decay with physical distance from patient and with time from COVID-19 diagnosis.
 We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects, and that increased COVID-19 disease severity would be associated with greater environmental contamination.
 
-* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital (i.e., hospital days 1,2,7,14,21,and 28).
+* __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28.
 Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
-Multiple surfaces were swabbed within each patient room using a Copan FLOQ(TM) swab and a sterile template (20cm2).
-Specific surfaces are listed in Table 1 with high-touch surfaces selected using structured observations prior to study start.
+Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm^2^ template to ensure consistent sampling.
+High-touch surfaces (**Table 1**) were selected using structured observations prior to study start.
 Swabs were immediately placed into buffered AVL and frozen prior to further processing.
-Specimens then underwent RNA extraction and RT-PCR targeting the SARS-CoV-2 N1 region in triplicate.
+Specimens then underwent RNA extraction (<span style="background-color:cyan">which kit?</span>) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.
 
@@
 
-* __Definition of Exposures and Outcomes__: 
-Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
-The primary study outcome of interest, detection of SARS-CoV-2 by PCR assay, was determined by detection of any quantifiable RNA.
+* __Definition of Exposures and Outcomes__: The primary outcome of interest, detection of SARS-CoV-2 by RT-qPCR assay, was determined by detection of any quantifiable RNA on <span style="background-color:cyan">any of three technical replicates</span>.
+The primary exposures of interest were (1) distance from the head of patient's bed, measured as above; (2) time from COVID-19 diagnosis and from the start of the local COVID-19 case wave; and (3) COVID-19 clinical severity.
+We categorized COVID-19 severity by the highest level of oxygen support required during the subjects' admission, wither no oxygen support required (mild disease), only nasal or facemask oxygen support required (moderate disease), significant oxygen support (i.e., high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation) required (severe disease).
 
@@
 We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms.
-We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA.
-We had hypothesized that the probability of RNA detection would decrease with distance.
-This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm].
-However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied.
-This is most likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch.
+We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surace (elevated or floor, high-touch or low-touch) did predict different levels of contamination.
+We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020].
+While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave.
+The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census.
+This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020].
+Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
+
+We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment.
+Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement.
+This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures.
+In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination.
+While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].
+
+Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm].
+Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased.
+The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch.
 We observed large differences in the probability of detecting surface contamination by surface type.
 Floors had high probability of contamination, which did not vary throughout the room.
-This is likely related to both the role of patient and healthcare personnel (HCP) foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room.
+This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room.
 This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020].
@@
 
-While we did not observe a measurable impact of distance on the probability of surface contamination with SARS-CoV-2 RNA, we did detect a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis.
-A similar finding was described by Chia et al [@Chia2020], where patients over 7 days from illness had lower probability of surface contamination within rooms.
-Interestingly, while the probability of floor contamination decreased with time from diagnosis, the probability remained relatively high at later time points.
-Combined with the finding that the probability of floor contamination increased with time from the onset of the second COVID-19 Wave, this suggests that external contamination accounts for a proportion of floor contamination.
-Viral RNA carried on the feet of HCP or the mops of EVS teams may have resulted in increased probability of detection as COVID-19 census increased.
-This is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020].
-Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
-
-With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms.
-This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures.
-In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination.
-Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].
-
 Our study has several limitations.
-First, our study used quantitative PCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
-Secondly, we did not control for cleaning methods on study units.
+First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
+Secondly, we did not measure differences in cleaning methods across different study units.
 However, cleaning practices did not undergo systematic changes during the study time period and sampling was performed routinely prior to daily cleaning.
-Finally, we did not perform matching sampling of occupying patients to explore the role of patient viral load and surface contamination.
+Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
 
@@
 However, surface type, severity of illness, and time predicted detection of surface contamination.
-Future studies are needed to determine the role of surface contamination in risk of infection among patients and HCP.
+Future studies should focus on how the observed surface contamination contributes to risk of infection among patients and HCP, and on how cleaning strategies can target surfaces found to be at high risk of contamination.
 

### 2021-09-28 manuscript.Rmd
Manuscript Revision

@@
 Matthew J.
-Ziegler, MD, MS (1,2,3); Selamawit Bekele (2,3); Sean Loughrey, MS (2,3); Emily Reesey, MS (2,2); Elizabeth Huang (2,3); Pam Tolomeo, MPH (2,3); Ebbing Lautenbach, MD, MPH, MS (1,2,3); Brendan J.
+Ziegler, MD, MS (1,2,3); Elizabeth Huang (2,3); Selamawit Bekele (2,3); Emily Reesey, MS (2,2); Pam Tolomeo, MPH (2,2), Sean Loughrey, MS (2,3); Michael Z.
+David, MD, PhD (1,2,3); Ebbing Lautenbach, MD, MPH, MS (1,2,3); Brendan J.
 Kelly, MD, MS (1,2,3) for the CDC Prevention Epicenters Program
@@
 MJZ - study design, data collection and analysis, manuscript;
+EH - specimen collection and processing;  
 SB - subject enrollment and specimen collection;  
@@
 ER - specimen processing and analysis;  
-EH - specimen collection and processing;   
 PT - subject enrollment and specimen collection, manuscript;  
@@
 
-* __Background__: SARS-CoV-2 is known to contaminate surfaces within healthcare settings, but the spatial and temporal extent of environmental contamination has not been precisely defined.
-We sought to define SARS-CoV-2 contamination of different surface types and how contamination changes over time.
+* __Background__: The spatial and temporal extent of SARS-CoV-2 environmental contamination has not been precisely defined.
+We sought to elucidate contamination of different surface types and how contamination changes over time.
 
-* __Methods__: We performed longitudinal sampling of surfaces within COVID-19 patient rooms while measuring the distance from each surface to the patient bed.
-We performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA and modeled the role of distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.
+* __Methods__: We sampled surfaces longitudinally within COVID-19 patient rooms, performed quantitative RT-PCR for the detection of SARS-CoV-2 RNA, and modeled distance, time, and severity of illness on the probability of detecting SARS-CoV-2 using a mixed-effects binomial model.
 
-* __Results__: The probability of detecting SARS-CoV-2 RNA did not vary with distance.
+* __Results__: The probability of detecting SARS-CoV-2 RNA in a patietn room did not vary with distance.
 However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
 The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
 
-* __Conclusions__: Distance from patient's bed did not predict SARS-CoV-2 RNA deposition in patient rooms, but surface type, severity of illness, and time predicted surface deposition.
+* __Conclusions__: Distance from patient’s bed did not predict SARS-CoV-2 RNA deposition in patient rooms, but surface type, severity of illness, and time from local pandemic wave predicted surface deposition.
 
@@
 
-Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the healthcare and hospital settings.
-In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed uninfected patients developed COVID-19.
+Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the hospital and other healthcare settings.
+In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previoulsy uninfected patients developed COVID-19.
 Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].
@@
 SARS-CoV-2 transmission can be mediated by droplets, aerosols, or fomites.
-Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv].
+Prior studies have demonstrated that SARS-CoV-2 extensively contaminates surfaces in the healthcare environment [@Morawska2021-mv; @Ong2021_1; @Ryu2020; @Moore2021; @Coil2021; @Redmond2021; @Ong2021_2].
 However, the spatial extent of contamination and how contamination changes during the course of hospitalization remain less well understood [@Mody2021-ag].
 
-We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021-vm].
+We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021].
 We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.
@@
 All rooms underwent daily cleaning, according to hospital protocol.
-Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface.
+Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface.
 Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.
@@
 Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
-Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm^2^ template to ensure consistent sampling.
-High-touch surfaces (**Table 1**) were selected using structured observations prior to study start.
+Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling.
+High-touch surfaces (Table 1) were selected using structured observations prior to study start.
 Swabs were immediately placed into buffered AVL and frozen prior to further processing.
-Specimens then underwent RNA extraction (<span style="background-color:cyan">which kit?</span>) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.
+Specimens then underwent RNA extraction (QIAmp Viral RNA Mini Kit) and quantitative reverse transcription polymerase chain reaction (RT-qPCR) (Thermo Quantstudio) with primers and probe targeting the SARS-CoV-2 N1 region (Centers for Disease Control and Prevention (CDC) RT-qPCR probe assay, IDT) in triplicate.
 
-* __Distance Data Collection__: Distance from the environmental sampling site to the patient's head of bed was recorded using a laser measuring device (Bosch GLM 20).
-Templates were affixed to each surface and the linear distance from the patient's head of bed to the sample site was measured.
+* __Distance Data Collection__: Distance from the environmental sampling site to the patient’s head of bed was recorded using a laser measuring device (Bosch GLM 20).
+Templates were affixed to each surface and the linear distance from the patient’s head of bed to the sample site was measured.
 
-* __Definition of Exposures and Outcomes__: The primary outcome of interest, detection of SARS-CoV-2 by RT-qPCR assay, was determined by detection of any quantifiable RNA on <span style="background-color:cyan">any of three technical replicates</span>.
-The primary exposures of interest were (1) distance from the head of patient's bed, measured as above; (2) time from COVID-19 diagnosis and from the start of the local COVID-19 case wave; and (3) COVID-19 clinical severity.
-We categorized COVID-19 severity by the highest level of oxygen support required during the subjects' admission, wither no oxygen support required (mild disease), only nasal or facemask oxygen support required (moderate disease), significant oxygen support (i.e., high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation) required (severe disease).
+* __Definition of Exposures and Outcomes__: The primary outcome of interest, detection of SARS-CoV-2 by RT-qPCR assay, was determined by detection of any quantifiable RNA on any of three technical replicates.
+Replicates were reviewed for inconsistencies and PCR was repeated in the event of disparate results.
+The primary exposures of interest were (1) distance from the head of patient’s bed, measured as above; (2) time from COVID-19 diagnosis and from the start of the local COVID-19 case wave; and (3) COVID-19 clinical severity.
+We categorized COVID-19 severity by the highest level of oxygen support required during the subjects’ admission: 1) no oxygen support required (mild disease); 2) only nasal or facemask oxygen support required (moderate disease); and 3) high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation (severe disease).
 
@@
 * __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces.
-There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021-vm], but the suface type was significantly associated with the probability of detection.
+There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection.
 Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces.
@@
 We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms.
-We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surace (elevated or floor, high-touch or low-touch) did predict different levels of contamination.
+We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces.
 We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020].
@@
 The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census.
-This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas [@Guo2020].
+This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020].
 Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
@@
 In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination.
-While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].
+While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
 
-Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm].
+Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021].
 Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased.
@@
 Floors had high probability of contamination, which did not vary throughout the room.
-This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room.
-This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020].
+This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room.
+This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020].
 In contrast to floors, walls had significantly lower probability of contamination.
@@
 Secondly, we did not measure differences in cleaning methods across different study units.
-However, cleaning practices did not undergo systematic changes during the study time period and sampling was performed routinely prior to daily cleaning.
+However, cleaning practices did not undergo systematic changes during the study period and sampling was performed routinely prior to daily cleaning.
 Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
@@
 In conclusion, SARS-CoV-2 RNA surface deposition did not vary with distance within patient rooms.
-However, surface type, severity of illness, and time predicted detection of surface contamination.
+However, surface type, severity of illness, and time from beginning of the local COVID-19 wave predicted detection of surface contamination.
 Future studies should focus on how the observed surface contamination contributes to risk of infection among patients and HCP, and on how cleaning strategies can target surfaces found to be at high risk of contamination.

### 2021-09-28 manuscript.Rmd
Manuscript Revision (minor addition)

@@
 This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020].
-Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
+Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.
 
@@
 In contrast to floors, walls had significantly lower probability of contamination.
-This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
+This may be due to the decreased settling of respiratory droplets or aeresols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
 

### 2021-09-28 manuscript.Rmd
Fix typos.

@@
 
-* __Results__: The probability of detecting SARS-CoV-2 RNA in a patietn room did not vary with distance.
+* __Results__: The probability of detecting SARS-CoV-2 RNA in a patient room did not vary with distance.
 However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
@@
 Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the hospital and other healthcare settings.
-In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previoulsy uninfected patients developed COVID-19.
+In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previously uninfected patients developed COVID-19.
 Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].
@@
 We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms.
-We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces.
+We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did.
+The highest probability of contamination was found with floors, followed by elevated high-touch surfaces.
 We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020].
@@
 Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement.
-This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures.
+This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory support.
 In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination.
-While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
+In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
 
@@
 This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room.
-This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020].
+Our findings match previous description of high probability of floor contamination relative to other surfaces[@Chia2020].
 In contrast to floors, walls had significantly lower probability of contamination.
-This may be due to the decreased settling of respiratory droplets or aeresols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
+This may be due to the decreased settling of respiratory droplets or aerosols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
 
@@
 First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
-Secondly, we did not measure differences in cleaning methods across different study units.
-However, cleaning practices did not undergo systematic changes during the study period and sampling was performed routinely prior to daily cleaning.
+Second, we did not measure differences in cleaning methods across different study units.
+However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning.
 Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

### 2021-12-08 manuscript.Rmd
Mark edits.

@@
 MJZ is supported by the National Institute for Allergy and Infectious Diseases (K23 AI143925).
+We thank the Bushman Laboratory at the University of Pennsylvania for quantitative PCR methods support.
 
@@
 ### Word Count:    
-?????
+- Abstract: 199  
+- Body: 2227
 
@@
 However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
+<mark>Increased surface contamination was observed in room where patients required high-flow oxygen, positive airway pressure, or mechanical ventilation</mark> (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+The probability of elevated surface contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
 
@@
 In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previously uninfected patients developed COVID-19.
-Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].
+Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing, <mark>with 91.7% of those exposed to such source patients developing infection</mark>[@Karan2021-cm].
 

### 2021-12-09 manuscript.Rmd
Revisions for review response.

@@
 Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability.
-All rooms underwent daily cleaning, according to hospital protocol.
+All rooms underwent daily cleaning, according to hospital protocol <mark>(**Supplemental Methods**).
 Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface.
-Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.
-Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
+Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.</mark> Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
 
@@
 * __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28.
-Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
+<mark>Sampling was performed each morning prior to daily cleaning.</mark> Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
 Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling.
@@
 Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces.
-High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces.
+High-touch elevated surfaces <mark>(mouse, keyboard, bed rail, and doorknob)</mark> had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces <mark>(walls)</mark>.
 **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
@@
 While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave.
-The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census.
+The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or <mark>other mobile medical equipment</mark> increases with the COVID-19 patient census.
 This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020].
@@
 In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination.
-In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].
+In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@Cdc2021-du].
 
@@
 This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room.
-Our findings match previous description of high probability of floor contamination relative to other surfaces[@Chia2020].
+Our findings match previous description of high probability of floor contamination relative to other surfaces, <mark>but the role of mops in the spread of viral RNA has not been previously described [@Chia2020; @Rutala2007-ki]</mark>.
 In contrast to floors, walls had significantly lower probability of contamination.
@@
 First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
-Second, we did not measure differences in cleaning methods across different study units.
+<mark>Few prior studies that have been successful in recovering infectious virus from patient rooms [@Lednicky2020-ue; @Santarpia2020-ll].</mark> Second, we did not measure differences in cleaning methods across different study units.
 However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning.
-Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
+<mark>Third, we were not able to differentiate contamination of floors that was attributable to healthcare personnel foot traffic as opposed to the settling of respiratory particles or body fluids attributable to the index patient.
+We did not observe large differences in the abundance of SARS-CoV-2 RNA across surfaces where SARS-CoV-2 RNA was detected.
+Therefore, our analysis focused exclusively on factors that are associated with the presence or absence of SARS-CoV-2, rather than SARS-CoV-2 abundance on positive surfaces.</mark> Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
 
@@
   gt::fmt_percent(columns = prop_positive, decimals = 1) %>%
+  #gt::tab_footnote()
   identity() -> dat_gt
@@
 
-__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are reproted.
+__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are <mark>reported.
+The "exit" descriptor indicates sites within the patient room near the exit or threshold; the "near" descriptor indicates sites within the patient room near the patient bed</mark>.
 

### 2021-12-09 manuscript.Rmd
Add roboto font.

@@
 output: html_document
-css: style_helvetica_double.css
+css: style_roboto_double.css
 ---

### 2021-12-09 manuscript.Rmd
Check references.

@@
 However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)).
-<mark>Increased surface contamination was observed in room where patients required high-flow oxygen, positive airway pressure, or mechanical ventilation</mark> (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
+<span style='background-color:lightcyan;'>Increased surface contamination was observed in room where patients required high-flow oxygen, positive airway pressure, or mechanical ventilation</span> (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)).
 The probability of elevated surface contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.
@@
 In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previously uninfected patients developed COVID-19.
-Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing, <mark>with 91.7% of those exposed to such source patients developing infection</mark>[@Karan2021-cm].
+Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing, <span style='background-color:lightcyan;'>with 91.7% of those exposed to such source patients developing infection</span>[@Karan2021-cm].
 
@@
 Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability.
-All rooms underwent daily cleaning, according to hospital protocol <mark>(**Supplemental Methods**).
+All rooms underwent daily cleaning, according to hospital protocol <span style='background-color:lightcyan;'>(**Supplemental Methods**).
 Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch surfaces, limiting use to a single surface.
-Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.</mark> Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
+Floors were cleaned using a microfiber mop and sodium dichloroisocyanurate, a chlorine-based disinfectant.</span> Patient rooms were eligible for inclusion in the study if the occupant had a positive test for SARS-CoV-2 within the prior 7 days and was admitted to a new patient room with COVID-19 isolation precautions.
 
@@
 * __Specimen Collection and SARS-CoV-2 RT-PCR__: COVID-19 patient rooms were swabbed longitudinally until day 28 or patient discharge from the hospital, with sampling targeted to hospital days 1, 2, 7, 14, 21, and 28.
-<mark>Sampling was performed each morning prior to daily cleaning.</mark> Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
+<span style='background-color:lightcyan;'>Sampling was performed each morning prior to daily cleaning.</span> Longitudinal sampling continued if the patient was removed from COVID-19 isolation precautions or the patient transferred to a new hospital room.
 Multiple surfaces were swabbed within each patient room using a flocked nylon swab (Copan) within a sterile 20 cm2 template to ensure consistent sampling.
@@
 Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces.
-High-touch elevated surfaces <mark>(mouse, keyboard, bed rail, and doorknob)</mark> had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces <mark>(walls)</mark>.
+High-touch elevated surfaces <span style='background-color:lightcyan;'>(mouse, keyboard, bed rail, and doorknob)</span> had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces <span style='background-color:lightcyan;'>(walls)</span>.
 **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.
@@
 While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave.
-The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or <mark>other mobile medical equipment</mark> increases with the COVID-19 patient census.
+The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or <span style='background-color:lightcyan;'>other mobile medical equipment</span> increases with the COVID-19 patient census.
 This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020].
@@
 This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room.
-Our findings match previous description of high probability of floor contamination relative to other surfaces, <mark>but the role of mops in the spread of viral RNA has not been previously described [@Chia2020; @Rutala2007-ki]</mark>.
+Our findings match previous description of high probability of floor contamination relative to other surfaces, <span style='background-color:lightcyan;'>but the role of mops in the spread of viral RNA has not been previously described [@Chia2020; @Rutala2007-ki]</span>.
 In contrast to floors, walls had significantly lower probability of contamination.
@@
 First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces.
-<mark>Few prior studies that have been successful in recovering infectious virus from patient rooms [@Lednicky2020-ue; @Santarpia2020-ll].</mark> Second, we did not measure differences in cleaning methods across different study units.
+<span style='background-color:lightcyan;'>Few prior studies that have been successful in recovering infectious virus from patient rooms [@Lednicky2020-ue; @Santarpia2020-ll].</span> Second, we did not measure differences in cleaning methods across different study units.
 However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning.
-<mark>Third, we were not able to differentiate contamination of floors that was attributable to healthcare personnel foot traffic as opposed to the settling of respiratory particles or body fluids attributable to the index patient.
+<span style='background-color:lightcyan;'>Third, we were not able to differentiate contamination of floors that was attributable to healthcare personnel foot traffic as opposed to the settling of respiratory particles or body fluids attributable to the index patient.
 We did not observe large differences in the abundance of SARS-CoV-2 RNA across surfaces where SARS-CoV-2 RNA was detected.
-Therefore, our analysis focused exclusively on factors that are associated with the presence or absence of SARS-CoV-2, rather than SARS-CoV-2 abundance on positive surfaces.</mark> Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
+Therefore, our analysis focused exclusively on factors that are associated with the presence or absence of SARS-CoV-2, rather than SARS-CoV-2 abundance on positive surfaces.</span> Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
 
@@
 
-__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are <mark>reported.
-The "exit" descriptor indicates sites within the patient room near the exit or threshold; the "near" descriptor indicates sites within the patient room near the patient bed</mark>.
+__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are <span style='background-color:lightcyan;'>reported.
+The "exit" descriptor indicates sites within the patient room near the exit or threshold; the "near" descriptor indicates sites within the patient room near the patient bed</span>.
 
